Formal models of biological networks in ANIMO 2

### 2013-12-09 Paper.tex
Writing about model checking, updating the comparison so that it is between 'ANIMO 1.0' and 'ANIMO 2.0'

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
The large amount of data with which an experimental biologist
must deal makes it largely impossible to reason on
the dynamics of complex systems without relying on computational
support.
In order to be fully profitable, computational modelling tools
must be both accessible and powerful, providing a means to
formalise and analyse experimental outcomes without requiring
additional training.
\parttitle{Results}
We show that the software tool ANIMO (Analysis of Networks with Interactive MOdelling) can be used
in different modelling tasks, and illustrate how this can be an asset for the biological researcher.
ANIMO models for \emph{Drosophila melanogaster} circadian clock and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells are used as case studies to illustrate the
possibilities offered by the tool.
\parttitle{Conclusions}
The software tool ANIMO, recently updated to version 2, allows the biologist to develop
a formal model through a user friendly interface,
achieving useful insight without the need to study any new mathematical formalisms.
\end{abstract}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
In living cells, processes are regulated by networks of interacting molecules.
Aberrations in these networks underlie a wide range of pathologies. The development of new therapies
requires a thorough insight in the functioning of these networks. Obtaining such insight can be a challenging task.
Feedback loops and crosstalk between pathways lead to an intricate wiring of the network.
Hence, it is necessary to study the ensemble of molecules involved, because the 
behaviour of individual molecules is not sufficient for a complete understanding. 
Since the human brain is ill-suited to grasp the non-linear dynamics of these complex networks and
the entailed emergent properties, the role of computational support is increasing
in molecular biology.

% The systems biology approach to understanding biological systems starts off from a
% scientific question and then follows an empirical cycle \--\ or rather a positive spiral \--\ of
% knowledge/theory $\rightarrow$ model $\rightarrow$ hypotheses $\rightarrow$ experiments $\rightarrow$
% observations $\rightarrow$ update and/or refinement of knowledge/theory,
% until an answer to the original question is found (Figure~\ref{fig:empirical-spiral}).
A model plays a pivotal role in this cycle:
\begin{enumerate}
  \item to organize data and store knowledge,
  \item to structure reasoning and discussion
  \item to perform \emph{in silico} experiments and derive hypotheses.
\end{enumerate}
An \emph{in silico} model is always a simplified representation of biological reality and is never the 
aim in itself.
% Rather, it is a powerful means in the process of gaining an understanding of the biological system.
% Given its role in the empirical cycle, the process of modelling is especially effective
% when applied by the experts with respect to a certain biological system. Biologists usually have a good sense 
% of cause-and-effect relationships of molecular interactions. In addition, they are the most knowledgeable
% on the network topology and the dynamics of the biological system they are studying.
% Since they also benefit most from the generation of hypotheses and from an efficient experimental design, 
% biologists would be the primary candidates to construct models of their research topic.

@@

\subsection*{\tas\ for biological models}
\tas\ have been developed to model the dynamic behaviour of systems with processes running in parallel~\cite{timed-automata-alur-dill}. 
As such, \tas\ have been applied in communication protocols and industrial control engineering~\cite{ta-audio-protocol,ta-wap-gateway,ta-ws-bap}. The 
parallels between these application areas and regulatory processes in cells have triggered the step towards 
their use in biology.
\cite{ta-siebert} use \tas\ to extend a classical modelling paradigm~\cite{thomas-formalism},
allowing to add temporal dynamics to gene network models.
\cite{bartocci-oscillators} describe a model of biological oscillators and test 
synchronization properties in this dynamic system.
A discretization of ODEs to \tas\ is proposed by~\cite{oded-ode-ta-discretization}, applying
a translation between the two formalisms to an example gene regulatory network. Two 
different approaches to transforming
a Petri net model into \tas\ are presented by~\cite{ta-giapponesi},
who also address the important issue of state space explosion in their paper.
Finally, \cite{ta-radiazioni} propose an \emph{ad hoc} \tas\ model of a radiation treatment
system, which is then validated through the analysis tool UPPAAL~\cite{uppaal}.\\
Each of these approaches has been successfully validated, demonstrating the potential of \tas\
in biological applications. However, these approaches were all limited to simple
or specific examples and none of these modelling methods
has led to a tool implementation of the proposed method to encourage a broader use
of \tas\ in molecular biology. We have recently introduced the software tool ANIMO
(Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee}), which aims
at making available the power of \tas\ to the experimental biologists without requiring
them to be fluent in any mathematical formalism.

@@

In this paper we present ANIMO 2.0, an updated version of ANIMO which allows for a considerably
more efficient analysis of models. ANIMO 2.0 also introduces new features, such as the possibility to interrogate
an ANIMO model with model checking queries represented as human language sentences instead of
mathematical formulae. The case studies presented in the following sections are centered around possible usage scenarios
for ANIMO, and aim at illustrating how a biologist can profit from the tool in its latest iteration.
An installation manual for ANIMO 2.0 is available at \url{http://fmt.cs.utwente.nl/tools/animo/manual.html}.

\subsection*{A brief introduction to ANIMO}
{\Large [[briefly describe ANIMO, its modelling assumptions and its intended use]]}

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation}\label{sec:animo-drosophila}
Results obtained with ANIMO are comparable to results with other modeling
approaches. To demonstrate this, Figure~\ref{fig:drosophila-model}
represents an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster}, based on the work
by~\cite{drosophila-ode-model}, where ordinary differential equations (ODEs) were used.
The cyclic behavior of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO act as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

\subsection*{Using ANIMO to generate hypotheses}\label{subsec:case-study-larger}
In order to validate our modeling approach,
we constructed a larger model of the signaling network downstream of TNF$\alpha$ and EGF, formalizing
the crosstalk that takes place between the pathways at different levels of cellular regulation.
We first modeled the two pathways in isolation~(Suppl. Figs.~\ref{fig:large-model-tnf}, \ref{fig:large-model-egf}),
using information on protein interactions from
the KEGG~\cite{kegg} and phosphosite~\cite{phosphosite} databases. These models were fitted to experimental data
from studies by~\cite{pathway-compendium} and~\cite{pathway-autocrine}.
We then merged the two pathways into a single model and added autocrine crosstalk between the pathways that has been
described by~\cite{pathway-autocrine}.
Briefly, stimulation with TNF$\alpha$ leads to a rapid release of TGF$\alpha$ ({\sf TGFa} in the model),
which activates the EGF receptor ({\sf EGFR}).
This activation causes secretion of IL-1$\alpha$ ({\sf IL-1a}) at later time points.
The effect of IL-1$\alpha$ is down-regulated by the secretion of IL-1 receptor antagonist ({\sf IL-1ra})
downstream of TNF$\alpha$.
The resulting model (Fig.~\ref{fig:large-model-no-hypotheses}) was compared to the experimental data
for treatments with 100 ng/ml TNF alone and 100 ng/ml EGF alone (data not shown)~\cite{pathway-compendium}.

At this point, the behavior of the model deviated from the data for some of the nodes.
This is an interesting situation, as it requires
modifications to the model, that can be interpreted as new hypotheses. Below, we give two examples and show how
adaptation of the model can be used to generate novel testable hypotheses.

@@

\subsection*{Testing the performances of ANIMO 2.0}\label{subsec:model-checking}
% \subsection*{Reaction-centered VS reactant-centered}\label{sec:animo-comparison}
% We will now apply some basic model checking queries to the case study presented in~\cite{animo-bibe},
% measuring the performances of the two modeling approaches. This will allow us to
% evaluate the benefit brought by the shift in perspective from a reaction- to a reactant-centered model.
In addition to some new features, the new ANIMO 2.0 provides a boost in terms of performances
(see Methods section for details).
In order to measure how significant the performance increase is, we applied a number of
ANIMO analysis queries to a case study, comparing the performances of ANIMO 2.0 with
the previous version of the tool. Moreover, ANIMO 2.0 provides the biolgist with
a proper access to the formal technique of model checking~\cite{model-checking}. As we will show, the
performance improvement is even more significant in this case, increasing
the potential for interactivity also in this area.

The case study we use as a testbed is the network model shown in the \emph{Network} panel in Figure~\ref{fig:cytoscape},
which represents signaling events downstream of growth factors EGF (epidermal growth factor)
and NGF (nerve growth factor) in PC12 cells (a cell line used to study neuronal differentiation).
The topology for this network was proposed in~\cite{egf-ngf}, and it was analyzed
with an ANIMO model based on the reaction-centered approach, reproducing the experimentally 
observed ERK (extracellular signal-regulated kinase) activity changes~\cite{animo-bibe}.
In particular, a 10 minutes stimulation with EGF resulted in transient behavior (i.e. peak-shaped,
see also the graph in Fig.~\ref{fig:cytoscape}), while NGF stimulation led to sustained activity (see Fig.~\ref{fig:comparison-graph-animo}).

\subsubsection*{Simulation-based analysis}
The most basic analysis that can be performed with ANIMO is based on simulating the behaviour
of the network during a given period of time. As this type of analysis is also normally
used when looking for a match between model parameters and experimental data, the computation
speed of a simulation run deeply affects the level of interactivity perceived by the user.
In order to make
the modeling approach in ANIMO as smooth as possible, it is desirable to decrease dead times,
i.e. significantly reduce simulation costs.
% UPPAAL is used in ANIMO to obtain simulation runs from a \tas\ model. While that process is
% very fast, 
% In this experiment, we query UPPAAL for simulation runs on the models
Our first experiment compares the execution time for simulation runs on the \tas\ models
generated by ANIMO applying the reaction- and reactant-centered modeling approaches to the case study.
To define the initial state of the model, we consider the starting condition to be the treatment with 50~ng/ml NGF, which translates into
setting the activity level of node NGF to 15/15, while changing EGF to be at 0/15 activity. We have chosen this configuration as the
model is expected to continue reacting indefinitely (sustained ERK activity), without reaching a deadlock state where all reactions are inactive.
This makes the model more demanding when generating the state space
because having more reactions occur means having more states in the transition system underlying the \tas\ model.
Table~\ref{tab:sim-100} illustrates the computation time and memory usage when performing 100 simulation runs on each of the
two considered models. Computing the simulation runs took about $92 \%$ less time with the reactant-centered model, using $87 \%$
less memory. This decrease in computation time for long simulation runs brings the approach nearer to the idea
of interactive exploration of a network.

@@

\subsection*{Model checking for bigger models}
When a model becomes particularly complex, with many events occurring concurrently,
the state space of the model tends to \emph{grow exponentially} in size, dramatically
increasing the time requested for the analysis of bigger models. To avoid the so-called
state space explosion, an approach that is becoming popular: the so-called \emph{statistical
model checking}.

@@

\begin{table}[h!]
\scriptsize
  \begin{center}
  \begin{tabular}{|l||r|r|}
    \hline
    Model type & Time (s) & Memory (peak KB) \\
    \hline
    \hline
    Reaction-centered & 37.14 & 257,928 \\
    \hline
    Reactant-centered & 2.83 & 33,464 \\
    \hline
  \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the query {\tt simulate 100 [36000] \{ R1, R2, \dots, R11 \}} on the model from~\cite{animo-bibe} with starting condition
  NGF = 15/15, EGF = 0/15, corresponding to a treatment with 50 ng/ml NGF.
  The query asks for 100 time series of the activity levels of all reactants in the model over the first 60 minutes
  of execution.\label{tab:sim-100}}
\end{table}

\begin{table*}[h!]
\scriptsize
  \begin{center}
    \begin{tabular}{|c||r|r||r|r||r|r|}
      \hline
       \multirow{3}{*}{Query} & \multicolumn{2}{c||}{Reaction-centered}	   & \multicolumn{2}{c||}{Reactant-centered} & \multicolumn{2}{c|}{Improvement}\\
      \cline{2-7}
       & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Time} & \multicolumn{1}{c|}{Memory} \\
       & \multicolumn{1}{c|}{time (s)}    & \multicolumn{1}{c||}{(peak KB)}    & \multicolumn{1}{c|}{time (s)} & \multicolumn{1}{c||}{(peak KB)} & \multicolumn{1}{c|}{(n-fold)} & \multicolumn{1}{c|}{(n-fold)}\\
      \hline
      \hline
      (1) & 127.248 & 351\,{}804 & 1.415 & 25\,{}716 & 90 & 14 \\
      \hline
      (2) & 167.635 & 269\,{}932 & 1.932 & 26\,{}112 & 87 & 10 \\
      \hline
      (3) & 150.901 & 318\,{}460 & 1.439 & 26\,{}136 & 105 & 12 \\
      \hline
      (4) & 2.154 & 216\,{}920 & 0.113 & 25\,{}316 & 19 & 9 \\
      \hline
      (5) & 470.560 & 311\,{}568 & 2.223 & 26\,{}648 & 212 & 12 \\
      \hline
      (6) & 453.900 & 357\,{}980 & 2.229 & 26\,{}652 & 204 & 13 \\
      \hline
    \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the given queries on the case study from~\cite{animo-bibe}.\label{tab:model-checking}}
\end{table*}

>>>>

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
The large amount of data with which an experimental biologist
must deal makes it largely impossible to reason on
the dynamics of complex systems without relying on computational
support.
In order to be fully profitable, computational modelling tools
must be both accessible and powerful, providing a means to
formalise and analyse experimental outcomes without requiring
additional training.
\parttitle{Results}
We show that the software tool ANIMO (Analysis of Networks with Interactive MOdelling) can be used
in different modelling tasks, and illustrate how this can be an asset for the biological researcher.
ANIMO models for \emph{Drosophila melanogaster} circadian clock and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells are used as case studies to illustrate the
possibilities offered by the tool.
\parttitle{Conclusions}
The software tool ANIMO, recently updated to version 2, allows the biologist to develop
a formal model through a user friendly interface,
achieving useful insight without the need to study any mathematical formalism.
\end{abstract}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
In living cells, processes are regulated by networks of interacting molecules.
Aberrations in these networks underlie a wide range of pathologies. The development of new therapies
requires a thorough insight in the functioning of these networks. Obtaining such insight can be a challenging task.
Feedback loops and crosstalk between pathways lead to an intricate wiring of the network.
% Hence, it is necessary to study the ensemble of molecules involved, because the 
% behaviour of individual molecules is not sufficient for a complete understanding. 
Since the human brain is ill-suited to grasp the non-linear dynamics of these complex networks and
the entailed emergent properties, the role of computational support is increasing
in molecular biology.
% 
% 
%
% The systems biology approach to understanding biological systems starts off from a
% scientific question and then follows an empirical cycle \--\ or rather a positive spiral \--\ of
% knowledge/theory $\rightarrow$ model $\rightarrow$ hypotheses $\rightarrow$ experiments $\rightarrow$
% observations $\rightarrow$ update and/or refinement of knowledge/theory,
% until an answer to the original question is found (Figure~\ref{fig:empirical-spiral}).
In this context, we can profit from models in different ways:
\begin{enumerate}
  \item to organize data and store knowledge,
  \item to structure reasoning and discussion
  \item to perform \emph{in silico} experiments and derive hypotheses.
\end{enumerate}
% An \emph{in silico} model is always a simplified representation of biological reality and is never the 
% aim in itself.
% Rather, it is a powerful means in the process of gaining an understanding of the biological system.
% Given its role in the empirical cycle, the process of modelling is especially effective
% when applied by the experts with respect to a certain biological system. Biologists usually have a good sense 
% of cause-and-effect relationships of molecular interactions. In addition, they are the most knowledgeable
% on the network topology and the dynamics of the biological system they are studying.
% Since they also benefit most from the generation of hypotheses and from an efficient experimental design, 
% biologists would be the primary candidates to construct models of their research topic.

@@

\subsection*{\tas\ for biological models}
\tas\ have been developed to model the dynamic behaviour of systems with processes running in parallel~\cite{timed-automata-alur-dill}. 
As such, \tas\ have been applied in communication protocols and industrial control engineering~\cite{ta-audio-protocol,ta-wap-gateway,ta-ws-bap}. The 
parallels between these application areas and regulatory processes in cells have triggered the step towards 
their use in biology.
In~\cite{ta-siebert}, the authors used \tas\ to extend a classical modelling paradigm~\cite{thomas-formalism},
allowing to add temporal dynamics to gene network models.
In~\cite{bartocci-oscillators}, a model of biological oscillators was described and synchronization properties
were tested in this dynamic system.
A discretization of ODEs to \tas\ was proposed in~\cite{oded-ode-ta-discretization}, applying
a translation between the two formalisms to an example gene regulatory network. Two 
different approaches to transforming
a Petri net model into \tas\ were presented in~\cite{ta-giapponesi},
where the important issue of state space explosion was also addressed.
Finally, the authors of~\cite{ta-radiazioni} proposed an \emph{ad hoc} \tas\ model of a radiation treatment
system, which was then validated through the analysis tool UPPAAL~\cite{uppaal}.\\
Each of these approaches has been successfully validated, demonstrating the potential of \tas\
in biological applications. However, these approaches were all limited to simple
or specific examples and none of these modelling methods
has led to a tool implementation of the proposed method to encourage a broader use
of \tas\ in molecular biology. We have recently introduced the software tool ANIMO
(Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee}), which aims
at making available the power of \tas\ to the experimental biologists without requiring
them to be fluent in any mathematical formalism.

@@

In this paper we present ANIMO 2.0, an updated version of ANIMO which allows for a considerably
more efficient analysis of models. ANIMO 2.0 also introduces new features, such as the possibility to interrogate
an ANIMO model with model checking queries represented as human language sentences instead of
mathematical formulae. The case studies presented in this paper are centered around possible usage scenarios
for ANIMO, and aim at illustrating how a biologist can profit from the tool in its latest iteration.
An installation and user manual for ANIMO 2.0 is available at \url{http://fmt.cs.utwente.nl/tools/animo/manual.html}.

\subsection*{A short introduction to ANIMO}
{\Large [[briefly describe ANIMO, its modelling assumptions and its intended use. Probably good idea
to refer to Gene paper]]}

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation}\label{sec:animo-drosophila}
Results obtained with ANIMO are comparable to results with other modeling
approaches. To demonstrate this, Figure~\ref{fig:drosophila-model}
represents an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster}, based on the work
by~\cite{drosophila-ode-model}, where ordinary differential equations (ODEs) were used.
The cyclic behavior of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

\subsection*{Using ANIMO to generate hypotheses}\label{subsec:case-study-larger}
We constructed a model of the signaling network downstream of TNF$\alpha$ and EGF, formalizing
the crosstalk that takes place between the pathways at different levels of cellular regulation.
We first modeled the two pathways in isolation~(Suppl. Figs.~\ref{fig:large-model-tnf}, \ref{fig:large-model-egf}),
using information on protein interactions from
the KEGG~\cite{kegg} and phosphosite~\cite{phosphosite} databases. These models were fitted to experimental data
from studies by~\cite{pathway-compendium} and~\cite{pathway-autocrine}.
We then merged the two pathways into a single model and added autocrine crosstalk between the pathways that has been
described by~\cite{pathway-autocrine}.
Briefly, stimulation with TNF$\alpha$ ({\sf TNFa} in the model) leads to a rapid release of TGF$\alpha$ ({\sf TGFa}),
which activates the EGF receptor ({\sf EGFR}).
This activation causes secretion of IL-1$\alpha$ ({\sf IL-1a}) at later time points.
The effect of IL-1$\alpha$ is down-regulated by the secretion of IL-1 receptor antagonist ({\sf IL-1ra})
downstream of TNF$\alpha$.
The resulting model (Fig.~\ref{fig:large-model-no-hypotheses}) was compared to the experimental data
for treatments with 100 ng/ml TNF alone and 100 ng/ml EGF alone (data not shown)~\cite{pathway-compendium}.

At this point, the behavior of the model deviated from the data for some of the nodes.
Changing the parameters of the model was not enough to reproduce the behaviour shown by experimental
data. This is an interesting situation, as it requires changes in the topology of
the model, which can be interpreted as new hypotheses. Below, we give two examples and show how
adaptation of the model can be used to generate novel testable hypotheses.

@@

\subsection*{Testing the performances of ANIMO 2.0}\label{subsec:model-checking}
% \subsection*{Reaction-centered VS reactant-centered}\label{sec:animo-comparison}
% We will now apply some basic model checking queries to the case study presented in~\cite{animo-bibe},
% measuring the performances of the two modeling approaches. This will allow us to
% evaluate the benefit brought by the shift in perspective from a reaction- to a reactant-centered model.
In addition to some new features, ANIMO 2.0 provides a boost in terms of performances
(see Methods section for details).
In order to measure how significant the performance increase is, we applied a number of
ANIMO analysis queries to a case study, comparing the performances of ANIMO 2.0 with
the previous version of the tool. Moreover, ANIMO 2.0 provides the biolgist with
a proper access to the formal technique of model checking~\cite{model-checking}. As we will show, the
performance improvement is even more significant in this case, making this type of formal
analysis more accessible.

The case study we use as a testbed is shown in Figure~\ref{fig:cytoscape},
and is a model signaling events downstream of growth factors EGF (epidermal growth factor)
and NGF (nerve growth factor) in PC12 cells.
The topology for this network was proposed in~\cite{egf-ngf}, and it was analyzed
with an ANIMO model, reproducing the experimentally 
observed ERK (extracellular signal-regulated kinase) activity changes~\cite{animo-bibe}.
In particular, a 10 minutes stimulation with EGF resulted in transient behavior (i.e. peak-shaped,
see also the graph in Fig.~\ref{fig:cytoscape}), while NGF stimulation led to sustained activity (see Fig.~\ref{fig:comparison-graph-animo}).

\subsubsection*{Simulation-based analysis}
The most basic analysis that can be performed with ANIMO is based on simulating the behaviour
of the network during a given period of time (see also the analyses
of the Drosphila and TNF-$\alpha$/EGF models above). As this type of analysis is also normally
used when looking for a match between model parameters and experimental data, the computation
speed of a simulation run deeply affects the level of interactivity perceived by the user.
In order to make
the modeling approach in ANIMO as smooth as possible, it is desirable to decrease dead times,
i.e. significantly reduce simulation costs.
% UPPAAL is used in ANIMO to obtain simulation runs from a \tas\ model. While that process is
% very fast, 
% In this experiment, we query UPPAAL for simulation runs on the models
Our first performance test compares the execution time for simulation runs on the \tas\ models
generated by ANIMO applying the old (i.e., ANIMO 1.0) and new (ANIMO 2.0) modeling approaches to the case study.
To define the initial state of the model, we consider the starting condition to be the treatment with 50~ng/ml NGF, which translates into
setting the initial activity level of nodes NGF to 15/15, and EGF to 0/15. We have chosen this configuration as the
model is expected to continue reacting indefinitely (sustained ERK activity), without reaching a deadlock state where all reactions are inactive.
This makes the model more demanding when generating the state space
because having more reactions occur means having more states in the transition system underlying the \tas\ model.
Table~\ref{tab:sim-100} illustrates the computation time and memory usage when performing 100 simulation runs on each of the
two considered models. Computing the simulation runs took about $92 \%$ less time with the new version of ANIMO, using $87 \%$
less memory. This decrease in computation time for long simulation runs brings the approach nearer to the idea
of interactive exploration of a network. This advantage is particularly noticeable when analysing
more complex networks.

@@

\subsection*{Model checking for bigger models}
When a model becomes particularly complex, and it represents many events occurring concurrently,
the state space of the model tends to grow exponentially in size, dramatically
increasing the time required for its analysis. To avoid this state space explosion,
\emph{statistical model checking} is commonly applied. This type of analysis is based on statistical
considerations and allows to obtain answers with a given confidence level in far less time
than most exact model checking techniques.

@@

\begin{table}[h!]
\scriptsize
  \begin{center}
  \begin{tabular}{|l||r|r|}
    \hline
    ANIMO Version & Time (s) & Memory (peak KB) \\
    \hline
    \hline
    ANIMO 1.0 & 37.14 & 257,928 \\
    \hline
    ANIMO 2.0 & 2.83 & 33,464 \\
    \hline
  \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the query {\tt simulate 100 [<=36000] \{ R1, R2, \dots, R11 \}} on the model from~\cite{animo-bibe} with starting condition
  NGF = 15/15, EGF = 0/15, corresponding to a treatment with 50 ng/ml NGF.
  The query asks for 100 time series of the activity levels of all reactants in the model over the first 60 minutes
  of execution. This means that, on average, it takes 0.37 seconds to perform one simulation of the behavior
  of the EGF/NGF model with ANIMO 1.0 and 0.028 seconds to compute the same result with ANIMO 2.0.\label{tab:sim-100}}
\end{table}

\begin{table*}[h!]
\scriptsize
  \begin{center}
    \begin{tabular}{|c||r|r||r|r||r|r|}
      \hline
       \multirow{3}{*}{Query} & \multicolumn{2}{c||}{ANIMO 1.0}	   & \multicolumn{2}{c||}{ANIMO 2.0} & \multicolumn{2}{c|}{Improvement}\\
      \cline{2-7}
       & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Time} & \multicolumn{1}{c|}{Memory} \\
       & \multicolumn{1}{c|}{time (s)}    & \multicolumn{1}{c||}{(peak KB)}    & \multicolumn{1}{c|}{time (s)} & \multicolumn{1}{c||}{(peak KB)} & \multicolumn{1}{c|}{(n-fold)} & \multicolumn{1}{c|}{(n-fold)}\\
      \hline
      \hline
      (1) & 127.248 & 351\,{}804 & 1.415 & 25\,{}716 & 90 & 14 \\
      \hline
      (2) & 167.635 & 269\,{}932 & 1.932 & 26\,{}112 & 87 & 10 \\
      \hline
      (3) & 150.901 & 318\,{}460 & 1.439 & 26\,{}136 & 105 & 12 \\
      \hline
      (4) & 2.154 & 216\,{}920 & 0.113 & 25\,{}316 & 19 & 9 \\
      \hline
      (5) & 470.560 & 311\,{}568 & 2.223 & 26\,{}648 & 212 & 12 \\
      \hline
      (6) & 453.900 & 357\,{}980 & 2.229 & 26\,{}652 & 204 & 13 \\
      \hline
    \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for ANIMO 1.0 and ANIMO 2.0 when computing
  the given queries on the case study from~\cite{animo-bibe}.\label{tab:model-checking}}
\end{table*}
